Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-06-16 paper/ifo-introduction.tex
a few more edits to the draft
working on paper comments from Andrew

% general gw intro
In 2015, gravitational waves were observed for the first time from the merger of two black holes in a binary system~\cite{GW150914}. 
The observation, made by the Laser Interferometer Gravitational-wave Observatory~\citep[LIGO]{AdvancedLIGO:2015}, marked a breakthrough in modern astrophysics and revealed a new means to observe the universe. 
Since 2015, the LIGO and Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
Gravitational-wave detection has led to increased public and media interest in gravitational-wave science. 
Gravitational-wave research groups around the world have produced demonstrations and activities to explain their research to general audiences. 
Activities range from hands-on demonstrations, exhibitions~\cite{TTExhibit:2020,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone apps~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}%\han{look up Sarah Farmer's project}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Albert Einstein's theory of general relativity~\cite{Einstein:1916}. 
They are disturbances in spacetime caused by the acceleration of massive objects. 
The effect of gravitational waves is a change in lengths: a `stretching and squeezing' of the distance between objects.
Observatories such as LIGO, Virgo, and KAGRA~\cite{KAGRA:2013} are laser interferometers, they use the interference of laser light to measure changes in distance. 
%As shown in an aerial view of LIGO Hanford in Figure~\ref{fig:ligo_pic}, they use beam arms kilometres long.
Gravitational-wave observatories are extremely complex, but are fundamentally are based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate lab experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

% continuous gws
To date the network of gravitational wave observatories have observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
However, the network is also searching for continuous gravitational waves; persistent periodic signals at near-monochromatic frequencies.
Continuous waves may be emitted at constant frequencies, or may wander slowly in frequency over time. 
Rotating neutron stars in low mass X-ray binaries (LMXBs) are one of the targets for continuous-wave searches. 
LMXBs are binary systems containing a compact object (such as a neutron star or a black hole) in orbit with a low mass stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is an LMXB which is particularly bright in X-rays. 
It is a prime target for continuous wave searches from LMXBs and numerous, as yet unsuccessful, searches have been performed~\cite{ScoX1O2Viterbi:2019,RadiometerO1O2:2019,SearchCrossCorrO1:2017}.

% sound analogy
We use the analogy of sound to demonstrate gravitational-wave science with a table-top experiment (see also Section~\ref{sec:ifo}). 
We use similar data analysis techniques to those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,SuvorovaEtAl:2017,SuvorovaEtAl:2017}.
As an undergraduate lab experiment this demonstration can be used to teach topics ranging from gravitational-wave detection and analysis to signal processing and speech enhancement. 
It can also be adapted for use as an outreach tool to explain gravitational-wave science to a general audience.

In this work we present a selection of demonstrations using different types of audio signals ranging from simple tones, analogous to continuous wave signals, to more complex audio of music and speech. 
In Section~\ref{sec:ifo} we detail our table-top interferometer design. 
We demonstrate observing a single note from a speaker in Section~\ref{sec:single_tone}.
In Section~\ref{sec:viterbi_wandering} we replace the single note with a slowly changing frequency and analyse the signal using techniques from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capturing and playing back complex audio, such as music and speech. 
This demonstration of an `optical microphone' serves as a more general exhibition of signal processing. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.

>>>>

% general gw intro
In 2015, gravitational waves were observed for the first time from the merger of two black holes~\cite{GW150914}. 
The observation, made by the Laser Interferometer Gravitational-wave Observatory~\citep[LIGO]{AdvancedLIGO:2015}, marked a breakthrough in modern astrophysics and revealed a new means to observe the universe. 
Since 2015, the LIGO and Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
Gravitational-wave detection has led to increased public and media interest in gravitational-wave science. 
Gravitational-wave research groups around the world have produced demonstrations and activities to explain their research to general audiences. 
Activities range from hands-on demonstrations, exhibitions~\cite{TTExhibit:2020,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone apps~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}%\han{look up Sarah Farmer's project}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Albert Einstein's theory of general relativity~\cite{Einstein:1916}. 
They are disturbances in spacetime caused by the acceleration of asymmetric massive objects. 
The effect of gravitational waves is a change in lengths: a `stretching and squeezing' of the distance between objects.
Observatories such as LIGO, Virgo, and KAGRA~\cite{KAGRA:2013} are laser interferometers.
They use the interference of laser light to measure changes in distance. 
%As shown in an aerial view of LIGO Hanford in Figure~\ref{fig:ligo_pic}, they use beam arms kilometres long.
Gravitational-wave observatories are extremely complex but are fundamentally based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate lab experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

% continuous gws
To date the network of gravitational wave observatories has observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals.
Continuous waves may be emitted at constant frequencies, or may wander slowly in frequency over time. 
For example, rotating neutron stars in low mass X-ray binaries (LMXBs) are one of the targets for continuous-wave searches. 
LMXBs are binary systems containing a compact object (such as a neutron star or a black hole) in orbit with a low mass stellar companion, where the rotation frequency of the compact object wanders over time due to variable accretion of matter (and hence angular momentum transfer) from the companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is an LMXB which is particularly bright in X-rays. 
It is a prime target for continuous wave searches from LMXBs and numerous, as yet unsuccessful, searches have been performed~\cite{ScoX1O2Viterbi:2019,RadiometerO1O2:2019,SearchCrossCorrO1:2017}.

% sound analogy
In this paper, we present an \han{inexpensive} table-top experiment, where a standard Michelson interferometer plays the role of a `toy' gravitational-wave detector, which detects acoustic vibrations (such as music) instead of gravitational waves. 
As an undergraduate lab experiment this demonstration can be used to teach topics ranging from gravitational-wave detection and analysis to electronics, signal processing, and speech enhancement. 
It gives students in courses such as physics and electrical engineering the freedom to explore the response of an accessible yet nontrivial optomechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,SuvorovaEtAl:2017,SuvorovaEtAl:2017}.
The demonstration can also be adapted for use as an outreach tool to explain gravitational-wave science to a general audience.

The paper presents 
In this work we present a selection of demonstrations using different types of audio signals ranging from simple tones, analogous to continuous wave signals, to more complex audio of music and speech. 
In Section~\ref{sec:ifo} we detail the table-top interferometer design. 
We demonstrate observing a single note from a speaker in Section~\ref{sec:single_tone}.
In Section~\ref{sec:viterbi_wandering} we replace the single note with a slowly changing frequency and analyse the signal using a hidden Markov model technique from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capturing and playing back complex audio, such as music and speech. 
This demonstration of an `optical microphone' serves as a more general exhibition of signal processing. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.
Appendix~\ref{app:code} collects the software and scripts used to produce this work.

### 2020-06-16 paper/paper_main.tex
a few more edits to the draft
working on paper comments from Andrew

@@

\begin{abstract}
Gravitational wave observatories around the world are searching for continuous waves: persistent signals from spinning neutron stars. 
Searches use sophisticated statistical techniques to look for weak signals in noisy data. 
We show how continuous wave searches can be demonstrated with a table-top model gravitational-wave detector (a Michelson interferometer). 
Sound is used as an analogue to gravitational-wave signals in our model detector. 
Using signal processing techniques from continuous wave searches we demonstrate the recovery of a constant and changing frequency tone. 
Finally we investigate the performance of the demonstration as an `optical microphone'. 
We test its recovery of music and speech. 
Simple chords and drums are easily recovered, however speech recordings are very noisy. 
Application of an existing speech enhancement technique provides some noise reduction, however the audio remains unintelligible. 
This demonstration has scope for use in the physics and electrical engineering undergraduate lab. 
It can also be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
\end{abstract}

>>>>

\author{Hannah Middleton}
\email{hannah.middleton@unimelb.edu.au}
\affiliation{%
 School of Physics, University of Melbourne, Parkville, Victoria, 3010, Australia
}
\affiliation{%
OzGrav-Melbourne, Australian Research Council Centre of Excellence for Gravitational Wave Discovery, Parkville, Victoria 3010, Australia
}

\author{Andrew Melatos}
\email{amelatos@unimelb.edu.au}
\affiliation{%
 School of Physics, University of Melbourne, Parkville, Victoria, 3010, Australia
}
\affiliation{%
OzGrav-Melbourne, Australian Research Council Centre of Excellence for Gravitational Wave Discovery, Parkville, Victoria 3010, Australia
}

\author{Robin Evans}
\email{robinje@unimelb.edu.au}
\affiliation{%
Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia
}
\affiliation{%
OzGrav-Melbourne, Australian Research Council Centre of Excellence for Gravitational Wave Discovery, Parkville, Victoria 3010, Australia
}

@@

\begin{abstract}
Gravitational wave observatories around the world are searching for continuous waves: persistent signals from spinning neutron stars. 
Searches use sophisticated statistical techniques to look for weak signals in noisy data. 
We show how continuous wave searches can be demonstrated with a table-top model gravitational-wave detector (a Michelson interferometer), where sound plays the role of gravitational waves. 
Using signal processing techniques from continuous wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
Finally we investigate the performance of the demonstration as an `optical microphone' for recovering music and speech. 
Simple chords and drums are easily recovered, but speech is more challenging. 
%however speech recordings are very noisy. 
%Application of an existing speech enhancement technique provides some noise reduction, however the audio remains unintelligible. 
This demonstration has scope for use in the physics and electrical engineering undergraduate lab. 
It can also be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
\end{abstract}
